Clustering based on probability distributions with application on residential customers

### 2021-10-19 main.Rmd
code smells less

```{r}
generate_design <- function(t, mu21, mu22, mu23){

t <- seq(0, t, 1)
g1 <- t %%2
g2 <- t %%3
g3 <- t %%5

# null design
g1_dnull <- rep( rep(0, each = length(unique(g1))), length.out= length(t))
g2_dnull <- rep( rep(0, each = length(unique(g2))), length.out= length(t))
g3_dnull <- rep( rep(0, each = length(unique(g3))), length.out= length(t))

# mean changing across categories in varying ways

g1_dvary <- rep(0, length.out= length(t))
g21_dvary <- rep(mu21, length.out= length(t))
g22_dvary <- rep(mu22, length.out= length(t))
g23_dvary <- rep(mu23, length.out= length(t))
g3_dvary <- rep(0, length.out= length(t))

design1 = distributional::dist_normal(g1_dnull + g2_dnull + g3_dnull)
design2 = distributional::dist_normal(g1_dnull + g21_dvary + g3_dnull)
design3 = distributional::dist_normal(g1_dnull + g22_dvary + g3_dnull)
design4 = distributional::dist_normal(g1_dnull + g23_dvary + g3_dnull)

data_bind <- tibble::tibble(
index = t,
g1 = g1,
g2 = g2,
g3 = g3,
design1 = distributional::generate(design1, times = 1) %>% unlist(),
design2 = distributional::generate(design2, times = 1) %>% unlist(),
design3 = distributional::generate(design3, times = 1) %>% unlist(),
design4 = distributional::generate(design4, times = 1) %>% unlist(),
#design5 = distributional::generate(design5, times = 1) %>% unlist()
) %>% 
  pivot_longer(-c(1, 2, 3, 4), names_to = "design", values_to = "sim_data")

data_bind
}

mu21= c(2, 0, 0)
mu22 = c(0, 2, 0)
mu23 = c(0, 0, 2)
mu3 = c(0, 0, 0, 0, 0)
data_bind <- generate_design(t, mu21, mu22, mu23)

```{r plot-gran, out.width="100%"}

p1 <- ggplot(data_bind,
             aes(x = index, y = sim_data)) + 
  geom_line() +
  xlab("index")+
  facet_wrap(~design, scales = "free_y",ncol =1) +
  theme_validation()

p2 <- ggplot(data_bind,
             aes(x = as.factor(g1), y = sim_data)) + 
  geom_boxplot(alpha =0.5) + xlab("g1") + 
  facet_wrap(~design, scales = "free_y", ncol = 1)+ stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 0.8, color = "blue") +
  theme_validation()

p3 <- ggplot(data_bind, aes(x = as.factor(g2), y = sim_data)) + geom_boxplot(alpha =0.5) + xlab("g2") + theme_bw() +
  facet_wrap(~design, scales = "free_y",ncol = 1)+ stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 0.8, color = "blue") + ylab("")+
  theme_validation()

p4 <- ggplot(data_bind, aes(x = as.factor(g3), y = sim_data)) + geom_boxplot(alpha =0.5) +
  xlab("g3") + theme_bw()+
  facet_wrap(~design, scales = "free_y", ncol = 1)+ stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 0.8, color = "blue")+ ylab("")+
  theme_validation()

gran2_change <- (p2 + p3 + p4) *
  theme_validation()
```

<!-- When we move from the linear to the cyclic world of temporal granularities, we can see patterns across different categories of the granularities which gets lost in the linear representation. Populations are modeled by a collection of these cyclic granularities. Each cyclic granularity might or might not have patterns across its categories. Each cluster is characterized by similar patterns across one or more of these cyclic granularities. -->

<!-- A small example is given to setup the problem. -->

<!-- We constructed the simulation parameters to represent common patterns in electricity data. 50, 200 and 500 time series were chosen for different simulation designs with different granularities and patterns changing across different granularities. -->
<!-- The data type is fixed to be "continuous".We generated independent replications of all combinations of the simulation parameters. -->

<!-- Consider a continuous time series variable $y$ of length $T$ indexed by ${0, 1, \dots T-1}$. Three circular granularities $g1$, $g2$ and $g3$ are considered with 2, 3 and 5 levels respectively. Categories of g1, g2 and g3 are represented by ${0,1}$, ${0, 1, 2}$ and ${0, 1, 2, 3, 4}$. These categories could be integers or some more meaningful labels. For example, the granularity "day-of-week" could be either represented by $\{0, 1, 2, \dots, 6\}$ or $\{Mon, Tue, \dots, Sun\}$. -->

<!-- Consider the case where distribution of $y$ would vary across levels of $g1$ and g2$ but will not vary across levels of $g3$ as described in Figure .Figure \ref shows four designs where $g3$ acts as a nuisance variable, and $g1$ and $g2$ varies in both  Design 2 and 3, and $g2$ varies in all designs. -->

<!-- Consider a case where distribution of $y$ would vary across levels of $g2$ -->
<!-- for all designs, across levels of $g1$ for few designs and $g3$ does not change across designs. Figure \ref{#fig:plot-linear} shows the linear and cyclic representation of $y$. The first panel shows raw plot of $y$ in a linear scale and the second panel shows distribution of $y$ across cyclic granularities namely $g1$, $g2$ and $g3$. As could be seen from the plots, it is impossible to decipher from the raw time plot that the time series variable shows such pattern across different granularities. -->

```{r change-var, echo=FALSE}
t = 300
n <- seq(0, t, 1)
g1 <- n %%2
g2 <- n %%3
g3 <- n %%5
mu11 = c(0, 5)
mu12 = c(3, 0)
mu21= c(2, 0, 0)
mu22 = c(0, 2, 0)
mu23 = c(0, 0, 2)
mu3 = c(0, 0, 0, 0, 0)
```

```{r generate-design} 
generate_design <- function(t, mu21, mu22, mu23){

t <- seq(0, t, 1)
g1 <- t %%2
g2 <- t %%3
g3 <- t %%5

# null design
g1_dnull <- rep( rep(0, each = length(unique(g1))), length.out= length(t))
g2_dnull <- rep( rep(0, each = length(unique(g2))), length.out= length(t))
g3_dnull <- rep( rep(0, each = length(unique(g3))), length.out= length(t))

# mean changing across categories in varying ways

g11_dvary <- rep(mu11, length.out= length(t))
g12_dvary <- rep(mu12, length.out= length(t))
g21_dvary <- rep(mu21, length.out= length(t))
g22_dvary <- rep(mu22, length.out= length(t))
g23_dvary <- rep(mu23, length.out= length(t))
g3_dvary <- rep(0, length.out= length(t))

design1 = distributional::dist_normal(g1_dnull + g2_dnull + g3_dnull)
design2 = distributional::dist_normal(g11_dvary + g21_dvary + g3_dnull)
design3 = distributional::dist_normal(g12_dvary + g22_dvary + g3_dnull)
design4 = distributional::dist_normal(g1_dnull + g23_dvary + g3_dnull)

data_bind <- tibble::tibble(
index = t,
g1 = g1,
g2 = g2,
g3 = g3,
design1 = distributional::generate(design1, times = 1) %>% unlist(),
design2 = distributional::generate(design2, times = 1) %>% unlist(),
design3 = distributional::generate(design3, times = 1) %>% unlist(),
design4 = distributional::generate(design4, times = 1) %>% unlist(),
#design5 = distributional::generate(design5, times = 1) %>% unlist()
) %>% 
  pivot_longer(-c(1, 2, 3, 4), names_to = "design", values_to = "sim_data")

data_bind
}

data_bind <- generate_design(t, mu21, mu22, mu23)

#data_bind

```{r plot-2gran, out.width="100%", fig.cap = "The linear (left) and cyclic (right) representation of the measured variable is shown. In this scenario, $g1$, $g2$ would change across atleast one design but $g3$ change remains same across all design. Thus $g3$ is not an important variable to distinguish these designs."}

# plot_linear_data <- function(data){
# ggplot(data,
#              aes(x = index, y = sim_data)) + 
#   geom_line() +
#   xlab("index")+
#   theme_bw() 
# }

p1 <- ggplot(data_bind,
             aes(x = index, y = sim_data)) + 
  geom_line() +
  xlab("index")+
  facet_wrap(~design, scales = "free_y",ncol =1) +
  theme_validation()

p2 <- ggplot(data_bind,
             aes(x = as.factor(g1), y = sim_data)) + 
  geom_boxplot(alpha =0.5) + xlab("g1") + 
  facet_wrap(~design, scales = "free_y", ncol = 1)+ stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 0.8, color = "blue") +
  theme_validation()

p3 <- ggplot(data_bind, aes(x = as.factor(g2), y = sim_data)) + geom_boxplot(alpha =0.5) + xlab("g2") + theme_bw() +
  facet_wrap(~design, scales = "free_y",ncol = 1)+ stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 0.8, color = "blue") + ylab("")+
  theme_validation()

p4 <- ggplot(data_bind, aes(x = as.factor(g3), y = sim_data)) + geom_boxplot(alpha =0.5) +
  xlab("g3") + theme_bw()+
  facet_wrap(~design, scales = "free_y", ncol = 1)+ stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 0.8, color = "blue")+ ylab("")+
  theme_validation()

gran1_change <- (p2 + p3 + p4) *
  theme_validation()
# 
# plot_cyclic_data( data_bind %>% filter(design=="design2"))
# plot_data( data_bind %>% filter(design=="design3"))
# plot_data( data_bind %>% filter(design=="design4"))
# plot_data( data_bind %>% filter(design=="design5"))
```

<!-- A subset of many possible designs are shown in Figure \ref{fig: plot-linear }. For the parameter space (XXX unique  combinations shown in table YYY), 100, 500 independent replications of all possible combination of simulation parameters were generated. The clustering methodologies were run all these unique combinations and subsets of these to verify if the methodologies work as expected. -->

@@

```{r gran2and1-clubbed, fig.cap=" For the left scenario $g1$, $g2$ would change across atleast one design but $g3$ change remains same across all design. For tthe right one, only $g3$ changes across different designs."}
#gran2_change  + gran1_change + plot_layout(widths = c(1,1))
#ggpubr::ggarrange(gran1_change, gran2_change, labels = c("a", "b"))
```

@@

```{r interaction-gran, fig.cap = "Design-1 has no change in distributions across different categories of $g1$ or $g2$, while Design-2 and Design-3 change across only $g1$ and $g2$ respectively. Design-4 changes across categories of both $g1$ and $g2$."}
## ----designs

sim_null_normal <- function(nxj, nfacetj, mean, sd, w1 = 0, w2=0) {
  rep(distributional::dist_normal(mu = mean, sigma = sd),
      times = nxj * nfacetj
  )
}

sim_varf_normal <- function(nx, nfacet, mean, sd, w1, w2) {
  rep(dist_normal((mean + seq(0, nfacet - 1, by = 1) * w1),
                  (sd + seq(0, nfacet - 1, by = 1) * w2)), each = nx)
}

sim_varx_normal <- function(nx, nfacet, mean, sd, w1, w2) {
  rep(dist_normal((mean + seq(0, nx - 1, by = 1) * w1),
                  (sd + seq(0, nx - 1, by = 1) * w2)), nfacet)
}

sim_varall_normal <- function(nx, nfacet, mean, sd, w1, w2) {
  dist_normal((mean + seq(0, (nx *nfacet - 1), by = 1) * w1),
              (sd + seq(0,  (nx *  nfacet - 1), by = 1) * w2))
}

## ----data

library(hakear)

nx_val = 2 # number of x-axis levels
nfacet_val = 3 # number of facet levels
w1_val = 1 # increment in mean
w2_val = 0 # increment in sd
mean_val = 0 # mean of normal distribution of starting combination
sd_val = 2 # sd of normal distribution of starting combination
quantile_prob_val = seq(0.1, 0.9, 0.1)
ntimes_val = 300

sim_panel_varall <- sim_panel(
  nx = nx_val, nfacet =  nfacet_val,
  ntimes = ntimes_val,
  # sim_dist = sim_varall_normal(2, 3, 5, 10, 5, -1.5)
  sim_dist = sim_varall_normal(nx_val, nfacet_val, mean_val, sd_val, w1_val, w2_val)
) %>% unnest(data)  %>% 
  rename( "g2" = "id_facet",
          "g1" = "id_x")

sim_panel_varx <- sim_panel(
  nx = nx_val, nfacet =  nfacet_val,
  ntimes = ntimes_val,
  # sim_dist = sim_varx_normal(2, 3, 5, 10, 5, -1.5)
  sim_dist = sim_varx_normal(nx_val, nfacet_val, mean_val, sd_val, w1_val, w2_val)
) %>% unnest(data)  %>% 
  rename( "g2" = "id_facet",
          "g1" = "id_x")

sim_panel_varf <- sim_panel(
  nx = nx_val, nfacet =  nfacet_val,
  ntimes = ntimes_val,
  # sim_dist = sim_varf_normal(2, 3, 5, 10, 5, 5)
  sim_dist = sim_varf_normal(nx_val, nfacet_val, mean_val, sd_val, w1_val, w2_val)
) %>% unnest(data)  %>% 
  rename( "g2" = "id_facet",
          "g1" = "id_x")

sim_panel_null <- sim_panel(
  nx = nx_val,
  nfacet =  nfacet_val,
  ntimes = ntimes_val,
  sim_dist = distributional
  ::dist_normal(mean_val, sd_val)
) %>% unnest(c(data)) %>% 
  rename( "g2" = "id_facet",
          "g1" = "id_x")

## ----category-plots

p_null <- sim_panel_null %>%
  ggplot(aes(x = as.factor(g1), y = sim_data)) +
  facet_wrap(~g2, labeller = "label_both") +
  geom_boxplot() +
  ylab("") +
  #ggtitle(paste("(a)", round(null, 2))) +
  xlab("") +
  theme_bw()+ stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 0.8, color = "blue") +
  #ggtitle("design1")+
  theme_validation() +ylab("")  +
  theme(panel.spacing =unit(0, "lines"))+ theme(
    strip.text = element_text(size = 8, margin = margin(b = 0, t = 0)))+
  theme(plot.margin = margin(0, 0, 0, 0, "cm") ) +ggtitle("Design-1")

p_varf <- sim_panel_varf %>%
  ggplot(aes(x = as.factor(g1), y = sim_data)) +
  facet_wrap(~g2, labeller = "label_both") +
  geom_boxplot() +
  ylab("") +
  #ggtitle(paste("(b)", round(varf, 2))) +
  xlab("")+
  theme_bw()+ stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 0.8, color = "blue") +
  #ggtitle("design2")+
  theme_validation()+ylab("") +
  theme(panel.spacing =unit(0, "lines"))+ theme(
    strip.text = element_text(size = 8, margin = margin(b = 0, t = 0)))+
  theme(plot.margin = margin(0, 0, 0, 0, "cm") )+ggtitle("Design-2")

p_varx <- sim_panel_varx %>%
  ggplot(aes(x = as.factor(g1), y = sim_data)) +
  facet_wrap(~g2, labeller = "label_both") +
  geom_boxplot()+
  ylab("")  +
  #ggtitle(paste("(c)", round(varx, 2))) +
  xlab("")+
  theme_bw()+ stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 0.8, color = "blue") +
  #ggtitle("design3")+
  theme_validation()+ylab("") +
  theme(panel.spacing =unit(0, "lines"))+ theme(
    strip.text = element_text(size = 8, margin = margin(b = 0, t = 0)))+
  theme(plot.margin = margin(0, 0, 0, 0, "cm") )+ggtitle("Design-3")

p_varall <- sim_panel_varall %>%
  ggplot(aes(x = as.factor(g1), y = sim_data)) +
  facet_wrap(~g2, labeller = "label_both") +
  geom_boxplot() +
  ylab("") +
  #ggtitle(paste("(d)", round(varall, 2))) +
  xlab("")+
  theme_bw()+ stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 0.8, color = "blue") +
  #ggtitle("design4") +
  theme_validation()+ylab("") +
  theme(panel.spacing =unit(0, "lines"))+ theme(
    strip.text = element_text(size = 8, margin = margin(b = 0, t = 0)))+
  theme(plot.margin = margin(0, 0, 0, 0, "cm") )+ggtitle("Design-4")

change_index <- function(data){

index_new <- map(seq_len(ntimes_val), function(i){
  map((seq_len(nx_val*nfacet_val)), function(j)
    {
value = i + (j-1)*ntimes_val
})
  }) %>% unlist()

data_new = data %>%
  ungroup() %>%
  mutate(index_old = row_number(),
         index_new = index_new)

y = data_new[match(index_new, data_new$index_old),]

y <- y %>%
  mutate(time = row_number())

return(y)
}

endbreaks<- nrow(sim_panel_null)

p1 <- change_index(sim_panel_null) %>%
  ggplot(aes(x = time,
             y = sim_data)) +
  geom_line() +
  scale_x_continuous(breaks = seq(1, endbreaks, 200))+
  theme_bw() +
  #geom_point(alpha = 0.5, color = "blue")
  #theme(panel.grid.major.x =  element_line(colour = "#A9A9A9"),
    #    panel.grid.minor.x =  element_blank())+
  ylab("")+
  xlab("")

p2 <- change_index(sim_panel_varf) %>%
  ggplot(aes(x = time,
             y = sim_data)) +
  geom_line() +
  scale_x_continuous(breaks = seq(1, endbreaks, 200))+
  theme_bw() +
  #geom_point(alpha = 0.5, color = "blue") +
  #theme(panel.grid.major.x =  element_line(colour = "#A9A9A9"),
     #   panel.grid.minor.x =  element_blank())+
  ylab("")+
  xlab("")

p3 <- change_index(sim_panel_varx) %>%
  ggplot(aes(x = time,
             y = sim_data)) +
  geom_line() +
  scale_x_continuous(breaks = seq(1, endbreaks, 200))+
  theme_bw() +
  #geom_point(alpha = 0.5, color = "blue")+
  #theme(panel.grid.major.x =  element_line(colour = "#A9A9A9"),
   #     panel.grid.minor.x =  element_blank())+
  ylab("")+
  xlab("")

p4 <- change_index(sim_panel_varall) %>%
  ggplot(aes(x = time,
             y = sim_data)) +
  geom_line() +
  scale_x_continuous(breaks = seq(1, endbreaks, 200))+
  theme_bw() +
  #geom_point(alpha = 0.5, color = "blue") +
  theme(panel.grid.major.x =  element_line(colour = "#A9A9A9"),
  panel.grid.minor.x =  element_blank())+
  ylab("")+
  xlab("")

#((p1 + p_null)/( p2 + p_varf)/(p3 + p_varx)/(p4 + p_varall)) 
#p_null+ p_varf + p_varx + p_varall
ggpubr::ggarrange(p_null, p_varf, p_varx, p_varall)

#labels = c("Design-1", "Design-2", "Design-3", "Design-4"))

>>>>

@@

```{r gran2and1-clubbed, fig.cap=" For the left scenario $g1$, $g2$ would change across atleast one design but $g3$ change remains same across all design. For the right one, only $g3$ changes across different designs."}
#gran2_change  + gran1_change + plot_layout(widths = c(1,1))
#ggpubr::ggarrange(gran1_change, gran2_change, labels = c("a", "b"))
```

@@

```{r interaction-gran, fig.cap = "Design-1 has no change in distributions across different categories of $g1$ or $g2$, while Design-2 and Design-3 change across only $g1$ and $g2$ respectively. Design-4 changes across categories of both $g1$ and $g2$."}
